---
title: Polarization in Congress
author: Jeff Lewis
description: Provides updated versions the classic Voteview plots of party mean DW-NOMINATE scores of overtime.  Figures and the underlying data are available for download. 
original_date: "2018-02-01" # Set the original date of your contribution in YYYY-MM-DD format here.
date: "`r format(Sys.time(), '%B %d, %Y')`" # Do not modify this
update_delta: 31
tags:
  - blog # Valid tags are: blog, help, data. You may use multiple tags for a post
  - data
---

```{r echo=FALSE, message=FALSE, error=FALSE}
# We use pacman to load and install dependencies.
library(pacman)

# Please list your article's dependencies here, comma separated:
p_load(tidyverse, ggplot2, ggrepel)
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here we create the polarization data and key graphics the Keith shows and describes in greater detail [here
](https://legacy.voteview.com/political_polarization_2015.htm).  You can download the `Rmarkdown` code that produces this file and all of the data sets and graphics presented [here](party_polarization.Rmd).

This polarization data set includes the following fields (again following Keith):

```
 0.  Chamber
 1.  Congress Number
 2.  First Year of the Congress
 3.  Difference in Party Means - first dimension
 4.  Proportion Moderates
 5.  Proportion of moderate Democrats (-0.25 to +0.25)
 6.  Proportion of moderate Republicans (-0.25 to +0.25)
 7.  Overlap
 8.  Chamber Mean - first dimension
 9.  Chamber Mean - second dimension
10.  Democratic Party Mean - first dimension
11.  Democratic Party Mean - second dimension
12.  Republican Party Mean - first dimension
13.  Republican Party Mean - second dimension
14.  Northern Republican Mean - first dimension
15.  Northern Republican Mean - second dimension
16.  Southern Republican Mean - first dimension
17.  Southern Republican Mean - second dimension
18.  Northern Democrat Mean - first dimension
19.  Northern Democrat Mean - second dimension
20.  Southern Democrat Mean - first dimension
21.  Southern Democrat Mean - second dimension
```

We begin by loading the data current nominate data from Voteview.com:
```{r message=FALSE}
nom_dat <- read_csv("https://voteview.com/static/data/out/members/HSall_members.csv")
```

To calculate the mean location of the Northern and Southern Democratic and Republican delegations, we
follow *Congressional Quarterly* in defining the "Southern states" as the 11 states of the Confederacy plus Oklahoma and Kentucky.  Note also the Democrats and Republicans are ICPSR party codes 100 and 200 respectively.

Here is the `R` code that transforms the member-year NOMINATE data into the chamber-year data on polarization:
 
```{r}
south <- c(40:49,51,53)
polar_dat <- nom_dat %>% 
    filter(congress>45 & 
           chamber != "President") %>%
    mutate( 
      year = 2*(congress-1) + 1789,
    ) %>%
    group_by(chamber,congress,year) %>% 
    summarize(
      party.mean.diff.d1 = mean(nominate_dim1[party_code==200],na.rm=T) - 
                           mean(nominate_dim1[party_code==100],na.rm=T),
      prop.moderate.d1 = mean(abs(nominate_dim1)<0.25,na.rm=T),
      prop.moderate.dem.d1 = mean(abs(nominate_dim1[party_code==100])<0.25,na.rm=T),
      prop.moderate.rep.d1 = mean(abs(nominate_dim1[party_code==200])<0.25,na.rm=T),
      overlap = (sum(nominate_dim1[party_code==200] <
                       max(nominate_dim1[party_code==100],na.rm=T),na.rm=T)  +
                 sum(nominate_dim1[party_code==100] >
                       min(nominate_dim1[party_code==200],na.rm=T),na.rm=T))/
                 (sum(!is.na(nominate_dim1[party_code==100]))+
                  sum(!is.na(nominate_dim1[party_code==200]))),
      chamber.mean.d1 = mean(nominate_dim1,na.rm=T),
      chamber.mean.d2 = mean(nominate_dim2,na.rm=T),
      dem.mean.d1 = mean(nominate_dim1[party_code==100],na.rm=T),
      dem.mean.d2 = mean(nominate_dim2[party_code==100],na.rm=T),
      rep.mean.d1 = mean(nominate_dim1[party_code==200],na.rm=T),
      rep.mean.d2 = mean(nominate_dim2[party_code==200],na.rm=T),
      north.rep.mean.d1 = mean(nominate_dim1[party_code==200 & 
                                             !(state_icpsr %in% south)],na.rm=T),    
      north.rep.mean.d2 = mean(nominate_dim2[party_code==200 & 
                                             !(state_icpsr %in% south)],na.rm=T),    
      south.rep.mean.d1 = mean(nominate_dim1[party_code==200 & 
                                              (state_icpsr %in% south)],na.rm=T),    
      south.rep.mean.d2 = mean(nominate_dim2[party_code==200 & 
                                             (state_icpsr %in% south)],na.rm=T),    
      north.dem.mean.d1 = mean(nominate_dim1[party_code==100 & 
                                              !(state_icpsr %in% south)],na.rm=T),    
      north.dem.mean.d2 = mean(nominate_dim2[party_code==100 & 
                                              !(state_icpsr %in% south)],na.rm=T),    
      south.dem.mean.d1 = mean(nominate_dim1[party_code==100 & 
                                              (state_icpsr %in% south)],na.rm=T),    
      south.dem.mean.d2 = mean(nominate_dim2[party_code==100 & 
                                              (state_icpsr %in% south)],na.rm=T),    
    ) 
```
Here are the first few lines of the resulting dataset:
```{r}
head(polar_dat)
write_csv(polar_dat,path="voteview_polarization_data.csv")
```
You can download this dataset as a csv [here](voteview_polarization_data.csv).


## Party means on the liberal-conservative dimension over time by chamber

Here are up to date versions of the famous party means over time plots for each chamber:
```{r echo=FALSE}
polar_dat_long <- polar_dat %>% gather(score,value,-chamber,-year,-congress)
labels <- c("dem.mean.d1"="DEM",
            "rep.mean.d1"="REP",
            "north.dem.mean.d1"="N. DEM",
            "south.dem.mean.d1"="S. DEM")

polarized_plot <- function(chamb) {
  pdatl <- polar_dat_long %>% 
                filter(chamber==chamb,
                       score %in% c("dem.mean.d1","rep.mean.d1",
                                    "north.dem.mean.d1","south.dem.mean.d1")) %>%
                mutate(party=labels[score]) %>%
                ungroup()
  
  gg <- ggplot(data=pdatl,
               aes(x=year,y=value,group=party,col=party)) +
               scale_x_continuous(expand = c(0.15, 0), 
                                 breaks=seq(1880, max(pdatl$year), by=8)) +
               geom_line() + geom_point(size=0.7) + 
               xlab("Year") + ylab("Liberal-Conservative") + 
               geom_text_repel(data=pdatl %>%
                    filter(year == min(year)),
                  aes(label = party, color = party),
                  size = 3,
                  nudge_x = -8,
                  point.padding = 0.1,
                  segment.color = NA,
                  show.legend = FALSE) +
              theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
              scale_color_manual(values=c("REP"="#c70828","DEM"="#1372ad",
                                         "N. DEM"="#6194F4",
                                         "S. DEM"="#81c4e4"),guide="none") +
              theme_bw()
  gg
}
```


### House
```{r echo=FALSE}
gg <- polarized_plot("House")
ggsave("voteview_house_party_means.png", plot=gg, device="png", 
       width=7, height=4, dpi=300)
ggsave("voteview_house_party_means.pdf", plot=gg, device="pdf", 
       width=7, height=4)
gg
```


You can download this graphic as a [png](voteview_house_party_means.png) or [pdf](voteview_house_party_means.pdf).

### Senate
```{r echo=FALSE}
gg <- polarized_plot("Senate")
ggsave("voteview_senate_party_means.png", plot=gg, device="png", 
       width=7, height=4, dpi=300)
ggsave("voteview_senate_party_means.pdf", plot=gg, device="pdf", 
       width=7, height=4)
gg

```


You can download this graphic as a [png](voteview_senate_party_means.png) or [pdf](voteview_senate_party_means.pdf).

## Liberal-conservative partisan polarization by chamber
```{r echo=FALSE}
gg <- ggplot(polar_dat_long %>% 
              filter(score == "party.mean.diff.d1"),
              aes(x=year,y=value,group=chamber,col=chamber)) +
              scale_x_continuous(expand = c(0.10, 0), breaks=seq(1880,max(polar_dat_long$year),by=8)) +
              geom_line() + geom_point(size=0.7) + 
              xlab("Year") + ylab("Distance between party means") + 
              geom_text_repel(data=polar_dat_long %>%
                    ungroup() %>%
                    filter(year == min(year) & score=="party.mean.diff.d1"),
                  aes(label = chamber, color = chamber),
                  size = 3,
                  nudge_x = -10,
                  point.padding = 0.1,
                  segment.color = NA,
                  show.legend = FALSE) +
              theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
              scale_color_manual(values=c("House"="#c70828","Senate"="#1372ad"),guide="none") +
              theme_bw() 

ggsave("voteview_party_mean_diff.png", plot=gg, device="png", 
       width=7, height=4, dpi=300)
ggsave("voteview_party_mean_diff.pdf", plot=gg, device="pdf", 
       width=7, height=4)
gg

```

You can download this graphic as a [png](voteview_party_mean_diff.png) or [pdf](voteview_party_mean_diff.pdf).

