
---
title: "Party Loyalty Scores"
author: "Jeff Lewis"
original_date: "2023-10-21"
date: "`r format(Sys.time(), '%B %d, %Y')`" # Do not modify this
output: html_document
update_delta: 31
description: Provides updated versions of Voteview calculations of party loyalty scores. Figures and the underlying data are available for download. 
tags:
  - blog # Valid tags are: blog, help, data. You may use multiple tags for a post
  - data
---

```{r echo=FALSE, message=FALSE, error=FALSE}
# We use pacman to load and install dependencies.
library(pacman)
# Please list your article's dependencies here, comma separated:
p_load(tidyverse, ggplot2)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Overview

Here we construct simple Party Loyality Scores for members of the House and Senate.   These data provide in tabular form the rates of voting with their party majority for each member. Members in parties with fewer then 10 members are dropped. 

You can download the loyalty scores in csv format [here](party_loyalty_scores.csv).

## Loading the roll call data

To begin, we download the current version of the Voteview roll call votes dataset as well as roll call and member meta data from Voteview:

```{r}
vote <- read_csv("https://voteview.com/static/data/out/votes/HSall_votes.csv",
                 show_col_types = FALSE)
member <- read_csv("https://voteview.com/static/data/out/members/HSall_members.csv",
                   show_col_types = FALSE) 
rc <- read_csv("https://voteview.com/static/data/out/rollcalls/HSall_rollcalls.csv",
               show_col_types = FALSE)
```

## Calculating the scores

We then calculate the majority choice for each party on each roll call and construct
the loyalty rates as well as congress-chamber floor and caucus median loyalty rates:

```{r warning = FALSE}
majority <- function(v) {
   m <- table(v)
   mv <- as.numeric(names(m)[which.max(m)])
   ifelse(max(m) > sum(!is.na(v))/2, mv, NA)
}

party_loyalty_score <- vote |> 
  left_join(rc |>
              select(congress, chamber, rollnumber, date), 
            by = c("congress", "chamber", "rollnumber")) |>
  select(-prob) |>
  left_join(member |> 
              select(chamber,
                     congress,
                     icpsr,
                     bioname, 
                     state_abbrev,
                     district_code,
                     party_code), 
            by = c("chamber", 
                   "congress",
                   "icpsr")) |>
  mutate(v = case_when(cast_code %in% 1:3 ~ 1, # add in paired and announced yeas
                       cast_code %in% 4:6 ~ 0, # add in paired and announced noes
                       TRUE ~ NA)) |>
  group_by(chamber, congress, rollnumber, party_code) |>
  mutate(party_majority = majority(v)) |>
  group_by(congress, chamber, icpsr, bioname, 
           state_abbrev, district_code, party_code) |>
  summarize(party_loyalty = 100*mean(v == party_majority, na.rm=TRUE),
            no_of_votes = sum(!is.na(v)),
           .groups = "drop")  |> 
  group_by(congress, chamber) |>
  mutate(floor_median_party_loyalty = median(party_loyalty, na.rm = TRUE)) |>
  group_by(congress, chamber, party_code) |>
  mutate(caucus_median_party_loyalty = median(party_loyalty, na.rm = TRUE),
         n_party = n()) |>
  ungroup() |>
  filter(n_party > 9)
```  

## Writing the data in csv form

Finally, we write the party loyalty dataset in csv format:

```{r}
max_cong <- max(party_loyalty_score$congress)
write_csv(party_loyalty_score |>
            filter(party_code %in% c(100, 200)) |>
            mutate(party = ifelse(party_code == 100, "D", "R")) |>
            select(congress:district_code, 
                   party, 
                   party_loyalty, 
                   no_of_votes, 
                   floor_median_party_loyalty,
                   caucus_median_party_loyalty),
            "party_loyalty_scores.csv")
```




