---
title: "Attendance of Senators and House Members Running for President"
author: "Jeff Lewis"
description: Track the rate of participation in roll call votes of Members of Congress who are running for President in 2020
original_date: "2019-07-30"
date: "`r format(Sys.time(), '%B %d, %Y')`"
tags:
  - blog
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pacman)
p_load(tidyverse, ggplot2, RColorBrewer)
color <- RColorBrewer::brewer.pal(3, "Set2")
```

The plot below shows the rates of participation in roll call votes by month for members of the House and Senate running for president.  For comparison, three colored lines are added to each panel.  The 
<font color="`r color[1]`">green line</font> shows the average rate of participation across the all of these presidential candidates.  The <font color="`r color[3]`">blue line</font> shows the average of Barack Obama's and Hillary Clinton's rates of participation in the corresponding months of the 2008 presidential election cycle.  The <font color="`r color[2]`">orange line</font> shows the average of Ted Cruz's and Marco Rubio's participation rates in the corresponding months of the 2016 presidential election cycle.   


```{r download data, message=F, warning=F, include=FALSE}
stack_csvs <- function(csvs) {
  lapply(csvs, read_csv, na=c("","NA","N/A")) %>%
     bind_rows()
}
votes <- stack_csvs(sprintf("https://voteview.com/static/data/out/votes/HS%i_votes.csv", 109:116)) 

mems <- stack_csvs(sprintf("https://voteview.com/static/data/out/members/HS%i_members.csv", 109:116))
          
rollcalls <- stack_csvs(sprintf("https://voteview.com/static/data/out/rollcalls/HS%i_rollcalls.csv", 109:116))
```

```{r make dataset, include=FALSE}
candidates <- tibble( icpsr = c(29147, 40700, 41301, 41308, 
                                20343, 20735, 41701, 21324, 40910),
                      cands = rep(TRUE, 9))

obamaClinton  <- tibble( icpsr = c(40520, 40105), 
                         oc = c(TRUE, TRUE) )

rubioCruz  <- tibble( icpsr = c(41304, 41102), 
                      rc = c(TRUE, TRUE) )



dat <- votes %>%
        left_join(mems, by=c("congress","icpsr","chamber")) %>% 
        left_join(rollcalls, by=c("congress","rollnumber","chamber")) %>%
        left_join(candidates, by=c("icpsr")) %>%
        left_join(obamaClinton, by=c("icpsr")) %>%
        left_join(rubioCruz, by=c("icpsr")) %>%
        mutate( oc = !is.na(oc), 
                cands = !is.na(cands),
                rc = !is.na(rc))
```

```{r echo=FALSE}
dat_by_month <- dat %>% 
                mutate(congress = ifelse(oc==1, congress+6, congress),
                       congress = ifelse(rc==1, congress+2, congress)) %>%
                filter(congress>114, (cands | oc | rc)) %>%
                mutate(date=as.Date(paste0(str_sub(date,1,7),"-01"))) %>%
                group_by(bioname, icpsr, date, congress, cands, oc, rc) %>%
                summarize(attend = sum(cast_code < 8)/n(),
                          n  = n())  %>%
                ungroup() %>%
                mutate(date = date + lubridate::years(ifelse(oc, 12, 0)))  %>%
                mutate(date = date + lubridate::years(ifelse(rc, 4, 0))) 
```

```{r echo=FALSE}

ggplot(dat_by_month %>% 
          filter(date > as.Date("2018-05-01"), cands==1), 
       aes(x=date, 
           y=100*attend)) +
  geom_line(data=dat_by_month %>% 
              filter(date > as.Date("2018-05-01"), cands) %>%
              group_by(date) %>%
              summarize(attend = mean(attend)),
            color=color[1], size=0.5, alpha=0.4) + 
  geom_line(data=dat_by_month %>% 
              filter(date > as.Date("2018-05-01"), oc,
                     date <= Sys.Date()) %>%
              group_by(date) %>%
              summarize(attend = mean(attend)),
            color=color[3], size=0.5, alpha=0.4) + 
  geom_line(data=dat_by_month %>% 
              filter(date > as.Date("2018-05-01"), rc,
                     date <= Sys.Date()) %>%
              group_by(date) %>%
              summarize(attend = mean(attend)),
            color=color[2], size=0.5, alpha=0.4) + 
  geom_line() + 
  geom_point() + 
  facet_wrap(~bioname, ncol=3) + 
  theme_minimal() +
  xlab("Month") + ylab("Attendence (Share of Roll Calls Cast)") +
  ylim(0,100) +
  scale_x_date(date_labels="%Y-%b") +
  theme(axis.text.x = element_text(angle = 55, hjust = 1)) 
```
